Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms

### 2023-10-06 manuscript/manuscript_2/paper.tex
introduction to P2

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022} is predicted to emit nHz gravitational waves (GWs). Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of nHz GWs has inspired the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost method is timing an ensemble of pulsars, i.e. a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}. A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth. By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA,][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}. These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep[CPTA,][]{Hobbs_2019}. \newline

The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. After multiple non-detections \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A} consilient evidence for the GW background was presented by NANOGrav \citep{2023ApJ...951L...8A}, EPTA/InPTA \citep{2023arXiv230616214A}, PPTA \citep{2023ApJ...951L...6R} and the CPTA \citep{2023RAA....23g5024X}. \newline

Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the early stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. Individual SMBHBs are continuous wave sources; they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals. Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs. However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}. \newline

Intrinsic pulsar timing noise -- i.e.  random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red). In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations. The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density. Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization. The new and existing approaches are therefore complementary. In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}. We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood (i.e. the model evidence) for model selection. \newline

% --- a promise which we aim to test as the key goal of this paper. 
\noindent This paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW. In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence. In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method. We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative example GW system. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space. Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively. The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization which is deferred to future work. Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

\section{State-Space Formulation}\label{sec:model}

In this Section we briefly review the state-space formulation of PTA analysis presented in K1.

@@

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed systems state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are identically correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column). The top panels show the true pulsar state $f_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$  (green curve). We plot the subtracted state to better illustrate the stochastic wandering of the pulsar frequency. In the left hand column, the blue/green solutions are completely overlaid. The middle panels show the true measured frequency $f_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve). Again we plot the subtracted state, and in the left hand column the red/magenta solutions are completely overlaid. The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{1}(t) - \hat{f}_{\rm m}^{1}(t)$. The results are shown for a single pulsar. In the right hand column $\Omega$ has been perturbed from its true value by 20\%} 
	\label{fig:kalman_example}
\end{figure*}

@@

\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
\begin{figure}
	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work. The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc. The grey dashed line denotes the Galactic plane.}
	\label{fig:pulsar_distrib}
\end{figure}
We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA. For each pulsar we adopt the fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist. The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$. The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits. The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value. Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$. In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
\end{eqnarray}
For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The vertical orange lines indicate the true injected values. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles. We are able to accurately estimate each parameter of interest. We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$. Note that the $x$-axis scaling does not span the total prior space.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\section{Exploring a broader parameter space} \label{sec:parameter_space}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_new}
	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}. Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$. The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals. Well estimated posteriors should fall along the diagonal $y=x$. $\Omega$ is generally well-estimated but many other parameters show evidence of being over-constrained due to a modelling bias. }
	\label{fig:parameter_space}
\end{figure}
We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}. In this section we test the method in different regions of the parameter space, varying the source parameters through astrophysically relevant ranges. \newline

@@

\subsection{Identifiability}\label{sec:identif}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
	\caption{Posterior probability distributions for the representative system described in Table \ref{tab:parameters_and_priors} over nine noise realisations for the GW source parameters $\iota$ and $h_0$ in the high SNR regime ($h_0 = 10^{-12}$). Unlike the other parameters of  $\boldsymbol{\theta}_{\rm gw}$, the one-dimensional posteriors for $\iota$ and $h_0$ do not strongly overlap. A correlation between  $\iota$ and $h_0$ is evident.}
	\label{fig:just_iota_h}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation. The red point labels the location of the likelihood maxima. The likelihood surface has been normalised with respect to the absolute value of the maxima. A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
	\label{fig:likelihood_surface}
\end{figure}
In Figure \ref{fig:just_iota_h} we plot the estimated one-dimensional posteriors for $\iota$ and $h_0$ for our high-SNR representative system, over nine noise realisations. 
The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but just for $\iota$ and $h_0$ rather than all members of $\boldsymbol{\theta}_{\rm gw}$. In the case of high-SNR we might expect \textit{a priori} that our estimates would be increasingly accurate over the low-SNR case, and that the method would converge robustly to similar posteriors irrespective of the noise realisation (i.e. $W_1 \to 0$). This is generally true for the majority of the component parameters of $\boldsymbol{\theta}_{\rm gw}$; the estimated posteriors across noise realisations are largely indistinguishable. However for $\iota$ and $h_0$ this is not the case. Instead the inferred one-dimensional posteriors do not broadly overlap and are not consistent across different noise realisations. Specifically across $10^3$ noise realisations the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$. For $\iota$, $W_{1,\rm median}$ is greater than the value inferred in the low SNR case ($=0.14$, Table \ref{tab:Wasserstein}), whilst for $h_0$ $W_{1,\rm median}$ is also greater as a proportion of the scale set by the strain amplitude. This is in contrast to $W_{1,\rm median}$ inferred for the other parameters of $\boldsymbol{\theta}_{\rm gw}$ in the high-SNR regime where as SNR increases, $W_1$ decreases; e.g. for $\Phi_0$, $W_{1, \rm median} = 4.5 \times 10^{-3}$ rad. \newline

This issue of inferring very different posteriors depending on the particular realisation of the noise is due to these parameters being only weakly-identifiable. Identifiability refers to whether it is theoretically possible to infer unique parameter values of the model, given the measured data and the model structure \citep{e5be7c83a0d24500826f6e1b414d1733}. Sources of non-identifiability are generally categorised as either ``structural", arising from the structure of the model, or ``practical", arising from insufficient data or measurement errors \citep{GUILLAUME2019418}. Regarding $\iota$ and $h_0$, we are concerned with structural identifiability. Indeed, one might initially suspect that $\iota$ and $h_0$ could have structural identifiability issues; from Equations \eqref{eq:hphx} and \eqref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$. For example, from Equation \eqref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$. This degeneracy is only broken by the relation for $h_{+}$, Equation \eqref{eq:hphx}. It can be shown (Appendix \ref{appendix_identifiability}) that $\iota$ and $h_0$ are in fact structurally identifiable, but with the caveat that they are only weakly-identifiable. That is to say the $\iota-h_0$ likelihood surface plateaus to a ridge with a small, non-zero gradient close to the maximum likelihood solution. This likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation. Whilst a single, unique maxima exists (labelled by the red point in the figure), along the ridge in the parameter space the likelihood values are all very similar. Consequently it is difficult for the nested sampling algorithm to select a particular point in this plateau and instead effectively trades off accuracy in one parameter against accuracy in other, depending on the particular realisation of the noise. If during the inference we fix one of the parameters at its true value (i.e. set a delta function prior), taking a slice through the likelihood surface it is then possible to identify a particular point in the plateau and the posteriors for the different noise realisations overlap indistinguishably, i.e. $W_1 \to 0$. In the low-SNR case these parameters are also weakly identifiable, but the increased measurement noise increases the uncertainty which effectively ``washes out" \footnote{\tiny \textcolor{red}{TK: "washes out" may be too colloquial / unspecific, but I think it is intuitive...}\normalsize}the likelihood ridge, leading to a more concave likelihood surface. This enables consistent posteriors to be estimated for $\iota$ and $h$. In the case where we fix e.g $h_0$ at its true value to infer indistinguishable posteriors for $\iota$, all of the posteriors for $\iota$ are all strongly biased away from the injected value; the maximum a posteriori probability estimates underestimate the injected value by $\sim 0.34$ rad. This is a more definitive sign of the bias that was suggested at in Sections \ref{sec:parameter_estim} and \ref{sec:multiple_noise} and will be now be discussed further in Section \ref{sec:bias}. 
\subsection{Bias by dropping pulsar terms}\label{sec:bias}

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
	\caption{As Figure \ref{fig:corner_plot_1} but for the high-SNR system with $h_0 = 10^{-12}$. For the majority of parameters the injection value does not fall within the 90\% credible interval and is not visible over these scales. The estimated posteriors are biased away from the injected value due to the dropping of the pulsar terms. Note that the scale of the $x$-axes are generally highly constrained compared to Figure \ref{fig:corner_plot_1}.} 
	\label{fig:bias_for_large_h}
\end{figure*}

>>>>

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) can be produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. A nHz GW influences the times of arrival (TOAs) of radio pulses from pulsars measured at the Earth. A pulsar timing array \citep[PTA][]{Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars. In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers. \newline

 PTAs search for two types of SMBHB GW signals. The first is a stochastic GW background due to the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Consilient evidence for the GW background has been presented by multiple separate PTAs \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X}, by measuring the cross-correlation between pairs of pulsars in the pulsar timing residuals as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. The second is resolved GWs from sufficiently close and massive individual SMBHBs \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. Individual SMBHBs are sources of continuous GWs. This results in persistent, quasi-monochromatic modulations in the pulsar timing residuals which can be modelled. Therefore, rather than cross-correlating pulsar pairs, individual SMBHBs are detected by either frequentist matched filtering \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or Bayesian inference \citep{Ellis2016,Arzoumanian2020A}. Observational efforts to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, although some inconclusive evidence at low significance has been presented \citep{2023arXiv230616226A}. \newline

 Irregularities in the intrinsic rotation of the pulsar result in random, unmodelled, red-spectrum TOA fluctuations known as `timing noise'. Timing noise limits the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. Additional white noise sources such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021} must also be considered. \cite{K1} (K23 hereafter) presented an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations. Specifically, K23 formulate PTA analysis as a state-space problem and use a Kalman filter \citep{Kalman1} to track the pulsars' intrinsic rotational state in conjunction with a Bayesian nested sampler \citep{Skilling, Ashton2022} for parameter estimation and model selection. \newline

 It is common in standard PTA analysis to drop the so called "pulsar term" from the modelling \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. Unlike the "Earth term", the pulsar term is not phase coherent and cannot be summed across the array to increase the total signal-to-noise. Instead it is considered as a source of self noise. However, dropping the pulsar terms results in a reduced detection probability ($\sim 5 \%$) bias in the parameter estimates \citep{Zhupulsarterms,Chen2022}. The method presented in K23 also drops the pulsar term. In this paper we remove this limitation and present the state-space approach to PTA data analysis including the pulsar term. \newline

 This paper is organised as follows. In Section \ref{sec:model} we briefly review the state-space formulation of PTA data analysis introduced in \cite{K1}. We do not review the methods for state tracking, parameter inference and model selection and refer the reader to K23. In Section \ref{sec:pulsar_term} we introduce how to include the pulsar term in the state-space model. In Section \ref{sec:rep_example1} we test the updated method on the synthetic data for a single representative example GW system, and compare the results to the original Earth-terms model. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space. Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively. Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

% Could include the below block if we need more fodder

%This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.

\section{State-Space Formulation}\label{sec:model}
In this section we briefly review the state-space formulation of PTA analysis presented in \citep{K1}.

@@

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}

A naive approach is challenging - look at this noisy likelihood curve

A reparamterisation is good

\section{Representative example: single SMBHB source and Comparison with Earth terms}\label{sec:rep_example1}

@@

@@

\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}

We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA. For each pulsar we adopt the fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist. The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$. The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits. The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value. Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$. In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
\end{eqnarray}
For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{example-image-a}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The vertical orange lines indicate the true injected values. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles. We are able to accurately estimate each parameter of interest. We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$. Note that the $x$-axis scaling does not span the total prior space.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\section{Exploring a broader parameter space} \label{sec:parameter_space}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{example-image-a}
	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}. Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$. The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals. Well estimated posteriors should fall along the diagonal $y=x$. $\Omega$ is generally well-estimated but many other parameters show evidence of being over-constrained due to a modelling bias. }
	\label{fig:parameter_space}
\end{figure}
We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}. In this section we test the method in different regions of the parameter space, varying the source parameters through astrophysically relevant ranges. \newline

@@

\subsection{Bias by dropping pulsar terms}\label{sec:bias}

Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g. the median of the one-dimensional marginalised posterior) is biased away from the true injection value. In Figure \ref{fig:bias_for_large_h} we show the resulting posteriors for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but now for the high-SNR system with $h_0 = 10^{-12}$. We can see that for all parameters, with the exception of $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it is not visible on the scales over which we have plotted the posteriors. This effect is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$. Similar results are obtained for other noise realisations. Whilst $h_0$ does not exhibit any noticeable bias for this particular noise realisation, in general across multiple noise realisations it does exhibit a correlated bias with $\iota$. In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}. \newline
